---
title: "Harjoitustyo"
author: "Ville Laitinen"
date: "8 5 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(tidyverse)
library(cowplot)
library(reshape2)
theme_set(theme_bw(20))
```


## Data


# 1. Likelihood and joint posterior density

```{r}

LL <- function(data, mu, lambda) {
  
  N <- length(data)
  
  
  ll <- log(mu[1])*(sum(data[1:lambda])) + 
    log(mu[2])*(sum(data[(lambda+1):N])) -
    lambda*mu[1] - (N-lambda)*mu[2]
  
  return(ll)
  
}


```

## Simulate data from the model

```{r}

N <- 200
lambda <- 100
mu <- c(5, 20)

sim_data <- c(rpois(lambda, mu[1]), rpois(N - lambda, mu[2]))


data.frame(y = sim_data, x = 1:N) %>%
  ggplot(aes(x = x, y = y)) +
  geom_vline(xintercept = lambda, color = "red", linetype = "dashed") +
  geom_point()
  

```


<!-- # ```{r} -->
<!-- # -->
<!-- # df <- lapply(1:25, function(i) { -->
<!-- #   lapply(1:25, function(j) { -->
<!-- # -->
<!-- #     LL(data, mu = c(i, j), 150) -->
<!-- # -->
<!-- #   }) %>% unlist %>% -->
<!-- #     cbind(value = ., mu1 = i, mu2 = 1:25) -->
<!-- # }) %>% do.call(rbind, .) -->
<!-- # -->
<!-- # -->
<!-- # df <- lapply(1:25, function(i) { -->
<!-- #   lapply(1:200, function(j) { -->
<!-- # -->
<!-- #     LL(data, mu = c(i, 15), j) -->
<!-- # -->
<!-- #   }) %>% unlist %>% -->
<!-- #     cbind(value = ., mu1 = i, mu2 = 1:200) -->
<!-- # }) %>% do.call(rbind, .) -->
<!-- # -->
<!-- # -->
<!-- # ``` -->
<!-- # -->




## Posterior 


<!-- ## Maximize likelihood -->

<!-- ```{r} -->

<!-- mle <- function(x) { -->

<!--   mu1 <- x[1] -->
<!--   mu2 <- x[2] -->
<!--   lambda <- x[3] -->

<!--   LL(sim_data, c(mu1, mu2), lambda) -->

<!-- } -->

<!-- optim(c(1, 1, 100), -->
<!--       mle, -->
<!--       lower = c(2, 2, 100), -->
<!--       upper = c(10, 50, 100), -->
<!--       method = "L-BFGS-B") -->

<!-- ``` -->



# 2. Full conditional densities

```{r}

r_mu1_cond <- function(data, mu2, lambda, alpha, beta) {
  
  shape <- sum(data[1:lambda]) + alpha[1] 
  rate <- lambda + beta[1]
  
  rgamma(1, shape = shape, rate = rate)

}

r_mu2_cond <- function(data, mu1, lambda, alpha, beta) {
  
  N <- length(data)
  
  # if lambda == last time points, there can be no change point
  # in this case return a sample from prior
  if(lambda == N) {
    return(rgamma(1, alpha[2], beta[2]))
  }
  
  shape <- sum(data[(lambda+1):length(data)]) + alpha[2] 
  rate <- N - lambda + beta[2]
  
  rgamma(1, shape = shape, rate = rate)
}

r_lambda_cond <- function(data, mu, alpha, beta) {

  N <- length(data)
  
  
  # Implement in log scale to avoid overflow
  c_log_mu <- sapply(1:N, FUN = function(i)  {
    
    (sum(data[1:i]) + alpha[1] - 1)*log(mu[1]) + ifelse(i == N, 0, (sum(data[(i+1):N]) + alpha[2] - 1)*log(mu[2]))
    
  }
  )
  
  
  unnormalized_log_density <- c_log_mu -
    (beta[1]*mu[1] + N*mu[2] + beta[2]*mu[2]) +
    (mu[2] - mu[1])*(1:N)  
  
  
  # Rescale and exp transform to normal scale
  density <- (unnormalized_log_density - 
    max(unnormalized_log_density) )%>% 
    exp
  
  # Normalize
  density <- density/sum(density)
  
  # Cumulative density
  cdf <- density %>% cumsum
  
  min(which(runif(1) <= cdf))
  
}

```


```{r}
max_lambda_cond <- function(data, mu, alpha, beta) {

  N <- length(data)
  
  
  # Implement in log scale to avoid overflow
  c_log_mu <- sapply(1:N, FUN = function(i)  {
    
    (sum(data[1:i]) + alpha[1] - 1)*log(mu[1]) + ifelse(i == N, 0, (sum(data[(i+1):N]) + alpha[2] - 1)*log(mu[2]))
    
  }
  )
  
  
  
  unnormalized_log_density <- c_log_mu -
    (beta[1]*mu[1] + N*mu[2] + beta[2]*mu[2]) +
    (mu[2] - mu[1])*(1:N)  
  
  log_density <- unnormalized_log_density/sum(unnormalized_log_density)
  
  which.max(log_density)
  
  
  
}



xx <- lapply(seq(from = 1, to = 100, length.out = 100), function(i) {
  lapply(seq(from = 1, to = 100, length.out = 100), function(j) {
    max_lambda_cond(data, c(i, j), alpha, beta)
    }) %>% unlist %>% cbind(j = 1:100, i = i)
  }) %>% do.call(rbind, .)


xx <- xx %>% set_colnames(c("max_lambda", "mu2", "mu1"))


p <- xx %>% as.data.frame() %>% ggplot(aes(x = mu1, y = mu2, fill = max_lambda)) + geom_tile()

p
print(p)

```



## Gibbs

```{r}

gibbs <- function(data, alpha, beta, inits = "random", n, burn_in = 0.25, keep = .5) {
  
  # Random initialization unless supplied in the function call
  if(inits == "random") {
    inits <- c(runif(2, 1, max(data)), runif(1, 1, N))
  }
  
  samples <- matrix(NA, n, 3) %>% 
    set_colnames(c("lambda", "mu1", "mu2"))
    
  samples[1, ] <- inits
  
  for(i in 2:n) {
    
    
    samples[i, "lambda"] <- r_lambda_cond(data = data,
                                          mu = samples[i-1, c("mu1", "mu2")],
                                          alpha = alpha,
                                          beta = beta)
    
    samples[i, "mu1"] <- r_mu1_cond(data = data,
                                    mu2 = samples[i-1, "mu2"],
                                    lambda = samples[i, "lambda"],
                                    alpha = alpha,
                                    beta = beta)
    
    samples[i, "mu2"] <- r_mu2_cond(data = data,
                                    mu1 = samples[i, "mu1"],
                                    lambda = samples[i, "lambda"],
                                    alpha = alpha,
                                    beta = beta)
    
  }
  
  # Remove burn-in
  samples <- samples[(round(nrow(samples)*burn_in) + 1):n, ]
  
  # thin
  samples <- samples[sample(1:nrow(samples), size = round(keep*nrow(samples))), ]
  
  return(samples)
  
}

```


# 3. Validate model with simulations

```{r}
# Hyperpriors
alpha <- c(2, 2)
beta <- c(0.25, 0.25)


N <- 200
lambda <- 100
mu <- c(5, 10)

sim_data <- c(rpois(lambda, mu[1]), rpois(N - lambda, mu[2]))


sim_samples <- gibbs(data = sim_data, 
      alpha = alpha, 
      beta = beta,
      inits = "random",
      n = 10000,
      burn_in = 0.5, 
      keep = 1)

sim_samples <- sim_samples %>% as.data.frame()

```

* The estimates are concentrated around the true values (see below). 

```{r, echo = FALSE}



p_lambda <- sim_samples %>% 
  ggplot(aes(x = lambda)) + 
  stat_density(geom = "line") + 
  geom_vline(xintercept = lambda, linetype = "dashed") + 
  labs(x = "Value", y = "Density", title = "Lambda")


p_mu1 <- sim_samples %>% 
  ggplot(aes(x = mu1)) + 
  stat_density(geom = "line") + 
  geom_vline(xintercept = mu[1], linetype = "dashed") + 
  labs(x = "Value", y = "Density", title = "Mu1")

p_mu2 <- sim_samples %>% 
  ggplot(aes(x = mu2)) + 
  stat_density(geom = "line") + 
  geom_vline(xintercept = mu[2], linetype = "dashed") + 
  labs(x = "Value", y = "Density", title = "Mu2")


p <- plot_grid(p_lambda, p_mu1, p_mu2, ncol = 1)

print(p)

```


* Below data with posterior mode and 75% quantiles for the inferred change point. 

```{r, echo = FALSE}
p <- data.frame(value = sim_data, x = 1:length(sim_data)) %>% 
  ggplot() + 
  geom_point(aes(y = value, x = x)) + 
  geom_vline(xintercept = which.max(sim_samples[, 1] %>% table) %>% 
               names %>%
               as.numeric(), color = "red") + 
  geom_vline(xintercept = quantile(sim_samples[, 1], probs = c(.25, .75)), color = "red", linetype = "dashed")
```



# 4. Real data

Let us know test our change point detector on data with unknown parameters.

```{r}

data <- c(4, 4, 3, 1, 3, 2, 1, 0, 11, 11, 12, 4, 4, 7, 9, 6, 9, 12, 13, 15, 12, 10, 10, 6, 6, 7, 12, 11, 
          15, 5, 11, 8, 11, 7, 11, 12, 14, 12, 8, 11, 9, 10, 6, 14, 14, 8, 4, 7, 10, 3, 14, 10, 17, 7,
          16, 9, 12, 11, 7, 11, 5, 11, 13, 9, 7, 9, 7, 11, 12, 13, 6, 9, 10, 13, 8, 18, 6, 16, 8, 4, 16, 
          8, 9, 5, 7, 9, 10, 11, 13, 12, 9, 11, 7, 9, 6, 7, 6, 11, 8, 5)

samples <- gibbs(data = data, 
      alpha = alpha, 
      beta = beta,
      inits = "random",
      n = 10000,
      burn_in = 0.5, 
      keep = 1)

```



* Plot estimates

```{r}
samples <- samples %>% as.data.frame()

p_lambda <- samples %>% 
  ggplot(aes(x = lambda)) + 
  stat_density(geom = "line") + 
  labs(x = "Value", y = "Density", title = "Lambda")


p_mu1 <- samples %>% 
  ggplot(aes(x = mu1)) + 
  stat_density(geom = "line") + 
  labs(x = "Value", y = "Density", title = "Mu1")

p_mu2 <- samples %>% 
  ggplot(aes(x = mu2)) + 
  stat_density(geom = "line") + 
  labs(x = "Value", y = "Density", title = "Mu2")


p <- plot_grid(p_lambda, p_mu1, p_mu2, ncol = 3)

print(p)
```



```{r}
samples %>% 
  as.data.frame() %>% 
  ggplot(aes(x = lambda)) + 
  stat_density(geom = "line")
  
```

