---
title: "demo7"
author: "Ville Laitinen"
date: "24 4 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(magrittr)
library(tidyverse)
library(mcmc)
theme_set(theme_bw(20))
```

# 1. Kumaraswamy

```{r}

# target distribution
kumaraswamy <- function(x, n, p) n*p*(1 - x^p)^(n-1)*x^(p-1)

# metropolis-hastings
kumaraswamy_samples <- function(N, init, n, p) {
  
  # initial value   
  y <- init
  
  for(i in 2:N) {
    
    
      proposal <- runif(1)
      alpha <- kumaraswamy(proposal, n, p)/kumaraswamy(y[i-1], n, p)
      
      if(runif(1) <= alpha) {
        y[i] <- proposal
      } else {
        y[i] <- y[i-1]
      }
     
  }
  
  y
}

```


* Sample and plot the Kumaraswamy distribution for some different n and p

```{r}

n_samples <- 1000

samples <- lapply(c(2, 3, 5, 10, 25), function(p) {
  
  lapply(c(5, 10, 50, 100), function(n) {
    data.frame(x = kumaraswamy_samples(n_samples, 0, n, p),
               n = rep(n, n_samples),
               p = rep(p, n_samples))
  }) %>%
    do.call(rbind, .)
  
}) %>% 
  do.call(rbind, .)

p <- samples %>% 
  ggplot(aes(x = x)) + 
  stat_density(geom = "line") +
  facet_grid(n ~ p, labeller = label_both)

```

```{r, echo=FALSE, fig.height=10, fig.width=15}
p
```

* Sample mean as a function of iterations with n = 100, p = 3

```{r}

mean_evol <- sapply(10*1:100, function(N) {
  
  c(m = kumaraswamy_samples(N, 0, n = 100, p = 3) %>% mean, iter = N)
  
}) %>% t %>% as.data.frame()

p <- mean_evol %>% 
  ggplot(aes(x = iter, y = m)) +
  geom_line()

p

```


* Chain autocorrelation

```{r}

some_samples <- kumaraswamy_samples(1000, 0, n = 100, p = 5)

autoc <- acf(some_samples)
```

* Autocorrelation is approximately 0 at x = 13. 

# 2. Slash  

```{r}

slash <- function(x) {
  
  if(x == 0) {
    y = 1/(2*sqrt(2*pi))
  } else {
    y = (dnorm(0) - dnorm(x)) / x^2
  }
  y
}
  
slash_samples <- function(N, init = 0) {
  
  y <- init
  
  for(i in 2:N) {
      proposal <- rnorm(1, y[i-1], 1)
      
      alpha <- slash(proposal)/slash(y[i-1])
      
      if(runif(1) <= alpha) {
        y[i] <-  proposal
      } else {
        y[i] <- y[i-1]
      }
  }
  y
}

```



## Plots

* Slash distribution

```{r}

ss <- slash_samples(1000)

data.frame(x = ss) %>% 
  ggplot(aes(x = x)) +
  stat_density(geom = "line")

```



```{r}

set.seed(11235)
mean_evol <- sapply(10*1:250, function(N) {
  
  c(m = slash_samples(N) %>% mean, iter = N)
  
}) %>% t %>% as.data.frame()

p <- mean_evol %>% 
  ggplot(aes(x = iter, y = m)) +
  geom_line()

print(p)
```

* There are weird extreme values of the mean. Lets' see what one of these outliers looks like:

```{r}


mean_evol <- lapply(10*1:100, function(N) {
  
  c(x = slash_samples(N), iter = N) 
  
})


ind <- lapply(mean_evol, mean) %>% unlist %>% which.max()

p <- as.data.frame(mean_evol[[ind]]) %>% 
  set_colnames("x") %>% 
  ggplot(aes(x = x)) +
  stat_density(geom = "line")

print(p)

```

* So the reason for a huge mean values is occasional outlier samples from the target.



# 3. mcmc::metrop

```{r}

# target log density
log_f <- function(x) -(x[1]^2 + x[2]^2 - 3)^2


n_samples <- 1000


samples <- lapply(c(0.01, 0.05, 0.1, 0.5, 1, 5), function(s) {
  xs <- metrop(log_f, c(0, 0), n_samples, scale = s)$batch
  
  xs <- xs %>%
    as.data.frame() %>% 
    set_colnames(c("x", "y")) %>% 
    cbind(scale = rep(s, n_samples))
  
}) %>% do.call(rbind, .)


```


* Plot

```{r}

p <- samples %>% 
  ggplot(aes(x = x, y = y)) +
  geom_line() + 
  geom_point(size = .5, color = "red") +
  facet_wrap(~scale) +
  labs(title = "MCMC paths with various scales")

print(p)

```

* Scale = 0.5 seems a decent value. 

* Try bad initial values

```{r}

inits <- list(c(5, 0), c(5, 5), c(5, 10), c(10, 10), c(5, -15), c(-20, 20))

samples <- lapply(1:length(inits), function(i) {
  xs <- metrop(log_f, initial = inits[[i]], n_samples, scale = 0.5)$batch
  
  xs <- xs %>%
    as.data.frame() %>% 
    set_colnames(c("x", "y")) %>% 
    cbind(inits = i)
  
}) %>% do.call(rbind, .)

p <- samples %>% 
  ggplot(aes(x = x, y = y)) +
  geom_line() + 
  facet_wrap(~inits, scales = "free") +
  labs(title = "MCMC paths with various initial values")

print(p)

```

