---
title: "demo2"
author: "Ville Laitinen"
date: "8 3 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(Matrix)
library(MASS)
library(magrittr)
library(tidyverse)
library(scatterplot3d)
library(gridExtra)

theme_set(theme_bw(20))
```



## 1. Sampling Distributions in R

* Define covariance matrix so that there's high correlation between 1. and 2. dimensions and variance of the first dimension is large compared to the other two: 

```{r}

Sigma <- matrix(c(10, 3, 0,
                  3, 1, 0,
                  0, 0, 1), 
                nrow = 3, ncol = 3)

# Sigma is positive definite:
nearPD(Sigma)$mat %>% as.matrix()

```


* Simulate observations from multivariate normal with the above covariance and mean 0:

```{r}

x <- mvrnorm(250, mu = c(0, 0, 0), Sigma = Sigma) %>% 
  as.data.frame()

```


* Pairwise plots:

```{r}
pairs(x)
```


* 3d plot: (plotly doesn't work, compatibility issues with rgl)

```{r}

p1 <- with(data = x,
     scatterplot3d(x = V1,
                   y = V2,
                   z = V3)
)

p2 <- with(data = x,
     scatterplot3d(x = V1,
                   y = V2,
                   z = V3, angle = -80)
)

```



## 2. Sampling an Arbitrary Distribution

### a) Sample the AADD with accept-reject

```{r}

# Function
arm_AADD <- function(n) {
  
  # parameters
  a <- 0
  b <- 3
  c <- 0.5
  
  # store samples
  samples <- c()
  
  while(length(samples) < n) {
    
    x <- runif(1, a, b)
    y <- runif(1, 0, c)
    
    if((x < 1 & y < .5*x) |
       (x > 2 & y < .5 - .5*(x - 2)) |
       (x < 2 & 1 < x & y < .5)) {
      samples <- c(samples, x)
    } 
    
  }
  
  return(samples)
}

# Sample and plot
p <- arm_AADD(5000) %>% 
  as.data.frame() %>% 
  ggplot(aes(x =  .)) +
  geom_histogram() +
  scale_x_continuous(limits = c(-1, 4))
  
p

```




### b) Sample the AADD with CDF

* First define the CDF: 

```{r}

AADD_CDF <- function(x) {
  
  if(x <= 0) {
    y <- 0
  } else if (x <= 1) {
    y <- .25*x^2
  } else if(x <= 2) {
    y <- .5*x - .25
  } else if(x <= 3) {
    y <- -.25*x^2 + 1.5*x - 1.25
  } else {
    y <- 1
  }
  
  return(y)
}


x <- seq(from = -1, to = 4, by = 0.01)
y <- sapply(x, AADD_CDF)

p <- data.frame(x, y) %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_point() +
  labs(subtitle = "CDF of AADD")

p
```

* Then the inverse: 

```{r}

inv_AADD_CDF <- function(x) {
  
  if(x < 0 | 1 < x) {
    stop("x not in [0, 1]")
  }
  
  if(x < .25) {
    y <- 2*sqrt(x)
  } else if(x < .75) {
    y <- 2*x + .5
  } else {
    y <- 3 - 2*sqrt(1-x)
  }
  
  return(y)
}


x <- seq(from = 0, to = 1, by = 0.01)
y <- sapply(x, inv_AADD_CDF)

p <- data.frame(x, y) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  labs(subtitle = "Inverse CDF of AADD")

p
```

* Inverse CDF seems ok. 

* Now sample form AADD with the inverse:

```{r}
p <- sapply(runif(5000, 0, 1), inv_AADD_CDF) %>% 
  as.data.frame() %>% 
  ggplot(aes(x =  .)) +
  geom_histogram() +
  scale_x_continuous(limits = c(-1, 4))
  
p

```

* Compare timing:

```{r}
inv_CDF_time <- system.time(sapply(runif(1000000, 0, 1), inv_AADD_CDF))

arm_time <- system.time(sapply(runif(1000000, 0, 1), arm_AADD))

arm_time
inv_CDF_time
```


## 3. Sampling from a Ball 

### a) Sampling rejection method

* Sample points from a 2d unit ball. 

```{r}

# Function
arm_2d_ball <- function(n) {
  
  n_samples <- rep(0, n)
  points <- matrix(0, n, 2)
  
  for(i in 1:n) {
    
    # Initialize
    x <- 2; y <- 2
    
    while(x^2 + y^2 > 1) {
      
      x <- runif(1, -1, 1)
      y <- runif(1, -1, 1)
      
      n_samples[i] <- n_samples[i] + 2
    }
    
    points[i, ] <- c(x, y)
    
  }
  
  
  return(list(points = points, n_samples = n_samples))
  
}

# Sample some points
x_2d_ball <- arm_2d_ball(10000)


p <- x_2d_ball$points %>%
  as.data.frame() %>%
  ggplot(aes(x = V1, y = V2)) +
  geom_point(size = .25) +
  labs(title = "Uniform(B_2)")

print(p)

# Samples needed on average (empirically)
mean(x_2d_ball$n_samples)

```


* Theoretically this is $$\text{Geom}(p)) = \text{Geom}(\frac{\pi}{4})) \to \frac{4}{\pi} \to \frac{8}{\pi} \approx 2.5464$$



### b)


### c) Sample from a 3D ball

* Clearly sampling a point from the ball requires only 3 samples from Unif(0, 1): 

```{r}

sample_3d_ball <- function(n) {
  points <- matrix(0, n, 3)

  for(i in 1:n) {
    # radius
    r <- runif(1, 0, 1)^(1/3)

    # 1. angle
    theta <- acos(1 - 2*runif(1, 0, 1))

    # 2. angle
    phi <- runif(1, 0, 1)*2*pi

    # Spherical coordinates to cartesian
    points[i, ] <- c(r*sin(theta)*cos(phi),
                     r*sin(theta)*sin(phi),
                     r*cos(theta))
    # points[i, ] <- c(r, theta, phi)
  }


  return(points)
}

# Sample some
x <- sample_3d_ball(1000)

# Check that all are inside B3
all(apply(x, 1, FUN = function(i) i^2 %>% sum) <= 1)

```


```{r}
# p <- with(data = x %>% as.data.frame(),
#      scatterplot3d(x = V1,
#                    y = V2,
#                    z = V3, angle = 0)
# )
# 
# print(p)

print(pairs(x %>% as.data.frame))

```

