---
title: "demo5"
author: "Ville Laitinen"
date: "2 4 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(magrittr)
library(tidyverse)
theme_set(theme_bw(20))
```


# 1. 

* Data
```{r}

df1 <- data.frame(yes = c(92, 161), no = c(108, 22)) %>% 
  set_rownames(c("A", "B"))

df1
```

* Odds ratio

```{r}

odds_ratio <- function(df) {
  
  r <- (df["B", "yes"]/df["B", "no"])/(df["A", "yes"]/df["A", "no"])
  
  r
}

OR = odds_ratio(df1)

OR

```

* Bootstrap

```{r}

group_A <- c(rep("yes", 92), rep("no", 108))
group_B <- c(rep("yes", 161), rep("no", 22))

boot_OR <- replicate( 
  rbind(
  sample(group_A, size = length(group_A), replace = TRUE) %>% 
  table,
  sample(group_B, size = length(group_B), replace = TRUE) %>% 
  table
  ) %>% 
  as.data.frame() %>% 
  set_rownames(c("A", "B")) %>% 
  odds_ratio(),
  n = 10000)

```


* 95% confidence interval

```{r}
quantile(boot_OR, probs = c(.025, .975))
```

* Bias

```{r}

mean(boot_OR - OR)

```

* Bootstrap slightly biased

* Plot estimate and CIs

```{r}

p <- ggplot() + 
  geom_histogram(aes(x = boot_OR)) +
  geom_vline(xintercept = OR) +
  geom_vline(xintercept = quantile(boot_OR, probs = c(.025, .975)), linetype = "dashed")

p

```


### Compare confidencde intervals

* Normal approximation 

```{r}

# Normal approximation
norm_mean <- boot_OR %>% mean
norm_sd <- boot_OR %>% sd

qnorm(p = c(.025, .975), norm_mean, norm_sd)

```

* Basic CI

```{r}
c(2*OR - quantile(boot_OR, probs = c((1 - (1 - .95))/2)),
  2*OR - quantile(boot_OR, probs = c((1 - .95))/2)) %>% 
  set_names(c("2.5%", "97.5%"))
```


* Bias corrected accelerated 

```{r}

# Edit df
long_df <- data.frame(group = c(rep("A", 200),
                     rep("B", 183)), answer = c(rep("yes", 92),
                                                rep("no", 108),
                                                rep("yes", 161),
                                                rep("no", 22)))

boot(long_df, statistic = function(df, ind) {
  odds_ratio(df[ind, ])
},
R = 1000
)

```



# 2. 

## a)

* Load data

```{r}
load("gas.RData")

head(gas)
```

* Scatter plot

```{r}

p <- gas %>% 
  ggplot(aes(x = EqRatio, y = NOx)) +  
  geom_point()

print(p)

```


* Fit splines

```{r}

splines <- smooth.spline(x = gas$EqRatio, y = gas$NOx, spar = 0.7)

p <- ggplot() + 
  geom_point(data = gas, aes(x = EqRatio, y = NOx)) + 
  geom_line(aes(x = splines$x, y = splines$y), color = "red")

p

```


* Bootstrap for uncertainty

```{r}


boot_splines <- lapply(1:1000, function(i) {
  
  df <- gas[sample(1:nrow(gas), nrow(gas), replace = TRUE), ]
    
  sp <- smooth.spline(x = df$EqRatio, y = df$NOx, spar = 0.7)
  
  data.frame(x = sp$x, y = sp$y, n = i)
}
) %>% 
  do.call(rbind, .)
 

p <- ggplot() + 
  geom_line(data = boot_splines, aes(x = x, y = y, group = n), alpha = 0.25, color = "chocolate") +
  geom_point(data = gas, aes(x = EqRatio, y = NOx), size = 1) +
  geom_line(aes(x = splines$x, y = splines$y))

p

```

## b) Parametric bootstrap

```{r}

std <- residuals(splines) %>% sd
mean <- residuals(splines) %>% mean 


# For each draw add to fit mean and compute splines
par_boot_splines <- lapply(1:1000, function(i) {
  
  res <- rnorm(length(splines$x), mean, std)
  
  boot_y <- splines$y + res
  
  sp <- smooth.spline(x = splines$x, y = boot_y, spar = 0.7)
  
  data.frame(x = sp$x, y = sp$y, n = i)
  
}) %>% 
  do.call(rbind, .)



p <- ggplot() + 
  geom_line(data = par_boot_splines, aes(x = x, y = y, group = n), alpha = 0.25, color = "chocolate") +
  geom_point(data = gas, aes(x = EqRatio, y = NOx), size = 1) +
  geom_line(aes(x = splines$x, y = splines$y))

p


```


* Parametric bootstrap produces smoother and more narrow confidence intervals. 