---
title: "demo4"
author: "Ville Laitinen"
date: "26 3 2019"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(magrittr)
library(reshape2)


```


# 1. Permutation testing
## 1.1 Plant growth; Mann-Whitney U-test

* Test groups

```{r}

data(PlantGrowth)

g1 <- subset(PlantGrowth, group == "ctrl")
g2 <- subset(PlantGrowth, group == "trt1")
g3 <- subset(PlantGrowth, group == "trt2")

```


* Define U-test function

```{r}

# test statistic
u_test <- function(x, y) {
  df <- outer(x, y, FUN = function(x, y) ifelse(x > y,
                                          1,
                                          ifelse(x == y, .5, 0))
  )
  
  sum(df)
}


# test
x <- rnorm(50, 0, 1)
y <- rnorm(25, 1, 2)
u_test(x, y) + u_test(y, x) == length(x)*length(y)

```


```{r}

# Null distribution
u_null <- function(x, y, n = 10) {
  
  sapply(1:n, FUN = function(i) {
    
    xy <- c(x, y)
    N <- length(xy)
    
    
    x_ind <- sample(1:N, length(x), replace =FALSE) 
    y_ind <- !(1:N %in% x_ind)
    
    rand_x <- xy[x_ind]
    rand_y <- xy[y_ind]
    
    u_test(rand_x, rand_y)
  })
  
}
  

mann_whitney <- function(x, y, two_tailed = FALSE,  null_n = 10000, null_dist = NULL) {
 
  if(is.null(null_dist)) {
    null_dist <- u_null(x, y, n = null_n)
  }
  
  p <- min(mean(u_test(x, y) > null_dist),
           mean(u_test(x, y) < null_dist))
  
  if(two_tailed) {
    p <- 2*p
  } 
  
  return(p)
}


```


* Compare to R's ```wilcox.test```

```{r}

true_values <- c(wilcox.test(g1$weight, g2$weight, alternative = "two.sided")$p.value,
                 wilcox.test(g1$weight, g3$weight, alternative = "two.sided")$p.value)


null_n <- 10000
null_distributions <- cbind(u_null(g1$weight, g2$weight, null_n),
                            u_null(g1$weight, g3$weight, null_n))

p_values <- sapply(1:null_n, FUN = function(i) {

  c(mann_whitney(g1$weight, g2$weight, two_tailed = TRUE,  null_dist = null_distributions[1:i, 1]),
  mann_whitney(g1$weight, g3$weight, two_tailed = TRUE, null_dist = null_distributions[1:i, 2]))
  
}) %>% 
  t %>% 
  as.data.frame() %>% 
  cbind(1:null_n) %>% 
  set_colnames(c("p1", "p2", "n"))

p <- p_values %>% 
  melt(id.vars = "n") %>% 
  ggplot(aes(x = n, y = value, color = variable)) +
  geom_line() +
  geom_hline(yintercept = true_values[1], linetype = "dashed", color = "#F8766D")+
  geom_hline(yintercept = true_values[2], linetype = "dashed", color = "#00A087B2")

print(p)

```


* Test with only 4 observations per group

```{r}

g1_4 <- sample(g1$weight, 4)
g2_4 <- sample(g2$weight, 4)
g3_4 <- sample(g3$weight, 4)

null_distributions2 <- cbind(u_null(g1_4, g2_4, null_n),
                            u_null(g1_4, g3_4, null_n))


# p-values
c(mann_whitney(g1_4, g2_4, null_dist = null_distributions2[, 1]),
  mann_whitney(g1_4, g3_4, null_dist = null_distributions2[, 2]))

```

* Null distribution histograms

```{r}

null_distributions2 %>% 
  as.data.frame() %>% 
  melt() %>% 
  ggplot(aes(x = value)) +
  geom_histogram(position = "dodge", binwidth = 1) +
  facet_wrap(~variable)

```


* Poor resolution.
 
* It is ok to compute two-tailed p = 2*one-tailed p, since we are interested in wether a treatment group grows better _or_ worse than control and because the distribution is symmetric.


 
## 1.2 Titanic; Chi squared-test



```{r}

# Edit data to pleasing shape
titan_table <- Titanic %>%
  as.data.frame() %>% 
  select(-Class) %>%
  group_by(Survived, Sex, Age) %>%
  summarize(sum = sum(Freq)) %>% 
  as.data.frame() %>% 
  ungroup()

titan_table


```

* Form Chi squared function

```{r}

# Chi squared statistic
chi_sq_statistic <- function(table) {
  
  if(dim(table)[1] != 2 & dim(table)[2] != 2) {
    stop("Invalid confingency table")
  }
  
  a <- table[1, 1]
  b <- table[1, 2]
  c <- table[2, 1]
  d <- table[2, 2]
  
  statistic <- ((a*d - b*c)^2)*(sum(table))/( (a+b)*(c+d)*(b+d)*(a+c) )
  
  return(statistic)
}

# Null distribution
chi_sq_null <- function(table, n = 100) {
  
  N <- sum(table)
  
  null_dist <- c()
  
  for(i in 1:n) {
    
    
    a <- sample(1:N, 1)
    b <- sample(1:(N - a), 1)
    c <- sample(1:(N - a - b), 1)
    d <- N - a - b - c
    
    null_dist <- c(null_dist,  chi_sq_statistic(matrix(c(a, b, c, d), 2, 2, byrow = TRUE)))
  }
  
  null_dist <- null_dist[is.finite(null_dist)]
  
  return(null_dist)
}

# Test against null
chi_sq_test <- function(table, null_dist = NULL, n = 100) {
  
  if(is.null(null_dist)) {
    null_dist <- chi_sq_null(table, n)
  }
  
  chi_sq <- chi_sq_statistic(table)
  
  p <- mean(null_dist >= chi_sq)
  
  return(p)

}

```


* Test effect of age on survival

```{r}
  age_df <- titan_table %>% 
  select(-Sex) %>%
  group_by(Age, Survived) %>%
  summarize(sum = sum(sum)) %>% 
  ungroup() %>% 
  spread(Age, sum) %>% 
  as.data.frame() %>%
  column_to_rownames(var="Survived")

chi_sq_test(age_df, n = 10000)

```
 
 
* Age seems independent of survival. Test effect of Sex.

```{r}
sex_df <- titan_table %>% 
  select(-Age) %>%
  group_by(Sex, Survived) %>%
  summarize(sum = sum(sum)) %>% 
  ungroup() %>% 
  spread(Sex, sum) %>% 
  as.data.frame() %>%
  column_to_rownames(var="Survived")

chi_sq_test(sex_df, n = 1000)

```

* Sex as well independent.
 
