---
title: "demo1"
author: "Ville Laitinen"
date: "5 3 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# Getting started

* Define variables:

```{r}
smallnum <- 4
bignum <- 124
```


* Check contents of environment:

```{r}
ls()
```


* Environment can be cleaned with ```rm(list=ls())```.


```{r}
# some basic operations
smallnum + bignum
bignum/smallnum
```

* In R code hash (#) is used for comments.   

* Sourcing doesn't print the commands to console resulting in a tidyer R experience.

* History tab contain all commands run in console.

## Vectors, matrices and data frames

* Define some vectors: 

```{r}
examplevector1 <- c(2, 4, 9, 5)
examplevector2 <- 1:25
examplevector3 <- seq(1,9,length=100)
examplevector4 <- seq(0,2*pi,by=pi/8)
```

* Access the first six elements of vector with ```head()```

```{r}
head(examplevector3)
```


* Define some matrices:

```{r}
examplematrix1 <- matrix(c(-1,2,3,1),nrow=2,ncol=2)
examplematrix2 <- matrix(1:12,nrow=3)
examplematrix3 <- t(examplematrix2)
```


* Matrices are stored as data, vectors as values in the environment.

* Elementwise matrix multiplication:

```{r}
examplematrix1*examplematrix1
```


* Matrix multiplication: 

```{r}
examplematrix1%*%examplematrix1
```

* Matrix multiplied by its inverse results in (approximate) identity matrix

```{r}
examplematrix1%*%solve(examplematrix1)
```


* Define data frame. Data frames are stored as data in the environment

```{r}
exampledataframe <- data.frame(Age = c(34,21,40,23), grade = LETTERS[1:4])
```




## Logical comparisons and loops

### Logical (boolean) variables

```{r}
logicalval <- FALSE
logicalval == TRUE # this is to test whether logicalval is equal to TRUE
logicalval == 1 # the same as the above
logicalval == 0 # indeed, logicalval is FALSE


smallnum < bignum
smallnum > bignum
! (smallnum < bignum)
smallnum + 120 == bignum


examplevector1 > smallnum
examplematrix3%*%examplematrix2 < bignum
```



### For and while loops


```{r}
for (k in 1:10) {
  # calculate whether smallnum is strictly less than k
  logicalval <- smallnum < k
  print(paste("It is", logicalval, "that", smallnum, "is strictly less than", k))
}

while(smallnum < 10){
  # while smallnum is strictly less than 10 add 1 to it
  print(paste(smallnum, "is strictly less than 10 so add 1 to it and test again..."))
  smallnum <- smallnum + 1
}
```

* A while loop is run as long as the condition is TRUE. 

### Functions

```{r}
if(bignum > 100){
  print("The number is strictly less than 100")
}else{
  print("The number is greater than or equal to 100")
}


strictinequalitycheck <- function(x){
  logicalval <- x < 100
  if(logicalval == 1){
    print("The number is strictly less than 100")
  }else{
    print("The number is greater than or equal to 100")
  }
}

strictinequalitycheck(smallnum)
strictinequalitycheck(bignum)
```

* Functions are stored as functions in the environment


# Exercise

## 1. 

```{r}
strictinequalitycheck <- function(x, test_against){
  logicalval <- x < test_against
  if(logicalval == 1){
    print(paste0("The number ", x," is strictly less than ", test_against))
  }else{
    print(paste0("The number ", x," is greater than or equal to ", test_against))
  }
}


strictinequalitycheck(bignum, 100)

```



## 2. Pseudorandom numbers

* Define random number generator: 

```{r}

rng <- function(X1, n, m, a, c) {
  
  # Check that parameters are ok
  if(!(0 < m &
       0 < a & a < m &
       0 <= c & c < m &
       1 <= n
  )) {
    stop("Check parameters!")
  }
  
  
  # Vector for output
  X <- c(X1)
  
  for(i in 2:n) {
    
    X_new <- (a*X[i-1] + c)%%m
    
    X <- c(X, X_new)
  }
  
  return(X)
}
```



* Test 

```{r}
random_numbers <- rng(1, 100, 64, 33, 4)
random_numbers
```


* Plot numbers

```{r}

p <- random_numbers %>% 
  data.frame(rnumber = ., index = 1:length(random_numbers)) %>% 
  ggplot(aes(x = index, y = rnumber)) +
  geom_point() +
  geom_line()
  
print(p)
```


* The numbers are clearly not random!

* Lets try with parameter c = 5:

```{r}
random_numbers <- rng(1, 100, 64, 33, 5)


p <- random_numbers %>% 
  data.frame(rnumber = ., index = 1:length(random_numbers)) %>% 
  ggplot(aes(x = index, y = rnumber)) +
  geom_point() +
  geom_line()
  
print(p)
```

* More randomness appears but still not impressively random. Now as c = 5 the cycle length becomes $2^6 = 64$.

* With even different parameters:

```{r}

random_numbers <- rng(1, 1000, 86436, 1093, 18257)


p <- random_numbers %>% 
  data.frame(rnumber = ., index = 1:length(random_numbers)) %>% 
  ggplot(aes(x = index, y = rnumber)) +
  geom_point() +
  geom_line()
  
print(p)


histogram <- random_numbers %>% 
  data.frame(rnumber = ., index = 1:length(random_numbers)) %>% 
  ggplot(aes(x = rnumber)) +
  geom_histogram()


histogram
```

* The trace plot looks very random but the histogram leaves room for doubt. Ideally with more generated numbers we would expect a more or less uniform distribution as all values are equally likely. This however is not enough as the order a histogram doesn't tell anything about the order in which the numbers were generated. 


* Let's try the generator with even more numbers: 
  
```{r}

random_numbers <- rng(1, 10000, 86436, 1093, 18257)


p <- random_numbers %>% 
  data.frame(rnumber = ., index = 1:length(random_numbers)) %>% 
  ggplot(aes(x = index, y = rnumber)) +
  geom_point() +
  geom_line()
  
print(p)


histogram <- random_numbers %>% 
  data.frame(rnumber = ., index = 1:length(random_numbers)) %>% 
  ggplot(aes(x = rnumber)) +
  geom_histogram(bins = 100)


histogram
```


* Very random indeed. 